A Comparison of Algorithms Playing EvoMan

### 2020-06-15 paper/paper.tex
PSO inertia weight updated set
Full table in tex

    The configurations used for the particle swarm optimization\cite{pso} algorithm are:
    \begin{itemize}
        \item population size: 30
        \item number of iterations: 200
        \item cognitive weight: \{0.4, 0.8, 1.5\}
        \item social weight: \{0.8, 0.4, 3\}
        \item inertia weight: \{constant 1, uniformly decreasing from 1 to 0.3\}
    \end{itemize}

@@

    \subsection{Random Initialization PPO vs PSO Cascading PPO}\label{subsec:random-initialization-ppo-vs-pso-cascading-ppo}
    The best PSO configuration was used in cascade before the PPO in 4 runs.
    PPO with random initialization was ran 3 times.
    The worst result of the PPO with random initialization is higher than the best result
    of PSO cascaded with PPO\@ (Fig.~\ref{fig:random_vs_pso_initialization}).
    \begin{figure}[htbp]
        \centering
        \includegraphics[width=0.5\textwidth]{images/random_vs_pso_initialization.eps}
        \caption{Gain comparison between PPO with random initialization and with PSO cascading.}
        \label{fig:random_vs_pso_initialization}
    \end{figure}
	Our explanation for getting worse results when using the technique mentioned above is that the function landscape is very big and PSO doesn't explore enough. Another possible explanation would be that the landscape is tricky when it comes to searching for generalist versus specialized solutions. We could try solving this problem by giving more exploratory power to PSO, but we did not experiment with this due to the lack of time. 
    \subsection{PPO on the best configuration}\label{subsec:ppo-on-the-best-configuration}
    Considering the results above, we decided that PPO with random initialization and enemies \{1, 2, 6, 7\} chosen for training would be the best configuration. We ran 3000 epochs of this configuration, saving all the models along the way after every 250 epochs.
    After taking a look at the testing results, we noticed that the final gain is greater after 2000 epochs than after 3000 epochs (Fig.~\ref{fig:ppo_2000_vs_3000_epochs}).
    This means that there is overfitting during the training.
    We can conclude that by stopping the PPO algorithm after 2000 epochs instead of 3000 epochs
    we can end up with more generalized agents that perform better on average against all opponents.
    \begin{figure}[htbp]
        \centering
        \includegraphics[width=0.5\textwidth]{images/ppo_2000_vs_3000_epochs.eps}
        \caption{Gain comparison between PPO after 2000 epochs and after 3000 epochs.}
        \label{fig:ppo_2000_vs_3000_epochs}
    \end{figure}

@@

    \subsection{Comparison with the Upper Bound}\label{subsec:comparison-with-the-upper-bound}
    With the PPO algorithm trained on the opponents \{1, 2, 6, 7\} we have
    beat the best specialized models from the original paper which were given as upper bounds\cite{evoman}.
    NEAT specialized
   \begin{table}[htbp]
        \caption{Specialized NEAT vs Generalized PPO on train opponents}
        \begin{center}
            \begin{tabular}{|c|c|c|}
                \hline
                \textbf{Opponent}&{\textbf{Specialized NEAT}}&{\textbf{Generalized PPO}} \\
                \cline{2-4}
                \hline
                1 & 190.01 & 199.81 \\
                2 & 194.01 & 190.34 \\
                6 & 173.01 & 195.85 \\
                7 & 177.01 & 193.27 \\
                \hline
                \multicolumn{3}{l}{The generalized PPO was trained on the opponents \{1, 2, 6, 7\}}
            \end{tabular}
        \end{center}
    \end{table}

@@

    \subsection{Time Analysis}\label{subsec:time-analysis}
    Since the match time is relevant, the machine on which the games were run is also relevant.
    The tests were run on an i7 4750HQ processor.
    The training was done on 3 threads.
    All computations were done on the CPU\@.
    The median game time during training is 28 seconds.
    (todo: include median nr of epochs at testing)

>>>>

    The configurations used for the particle swarm optimization\cite{pso} algorithm are:
    \begin{itemize}
        \item population size: 30
        \item number of iterations: 200
        \item cognitive weight: \{0.4, 0.8, 1.5\}
        \item social weight: \{0.8, 0.4, 3\}
        \item inertia weight: \{constant 1, decreasing from 1 by 0.0035 every epoch\}
    \end{itemize}

@@

    \subsection{Random Initialization PPO vs PSO Cascading PPO}\label{subsec:random-initialization-ppo-vs-pso-cascading-ppo}
    The best PSO configuration was used in cascade before the PPO in 4 runs.
    PPO with random initialization was ran 3 times.
    The worst result of the PPO with random initialization is higher than the best result
    of PSO cascaded with PPO\@ (Fig.~\ref{fig:random_vs_pso_initialization}).
    \begin{figure}[htbp]
        \centering
        \includegraphics[width=0.5\textwidth]{images/random_vs_pso_initialization.eps}
        \caption{Gain comparison between PPO with random initialization and with PSO cascading.}
        \label{fig:random_vs_pso_initialization}
    \end{figure}
    Our explanation for getting worse results when using the technique mentioned above is that the function landscape is very big and PSO doesn't explore enough. Another possible explanation would be that the landscape is tricky when it comes to searching for generalist versus specialized solutions. We could try solving this problem by giving more exploratory power to PSO, but we did not experiment with this due to the lack of time.

    \subsection{PPO on the best configuration}\label{subsec:ppo-on-the-best-configuration}
    Considering the results above, we decided that PPO with random initialization and enemies \{1, 2, 6, 7\} chosen for training would be the best configuration. We ran 3000 epochs of this configuration, saving all the models along the way after every 250 epochs.
    After taking a look at the testing results, we noticed that the final gain is greater after 2000 epochs than after 3000 epochs (Fig.~\ref{fig:ppo_2000_vs_3000_epochs}).
    This means that there is overfitting during the training.
    We can conclude that by stopping the PPO algorithm after 2000 epochs instead of 3000 epochs
    we can end up with more generalized agents that perform better on average against all opponents.
    \begin{figure}[htbp]
        \centering
        \includegraphics[width=0.5\textwidth]{images/ppo_2000_vs_3000_epochs.eps}
        \caption{Gain comparison between PPO after 2000 epochs and after 3000 epochs.}
        \label{fig:ppo_2000_vs_3000_epochs}
    \end{figure}

@@

    \subsection{Comparison with the Upper Bound}\label{subsec:comparison-with-the-upper-bound}
    With the PPO algorithm trained on the opponents \{1, 2, 6, 7\} we have
    beat the best specialized models from the original paper which were given as upper bounds\cite{evoman}.
    \begin{table}[htbp]
        \caption{Specialized NEAT vs Generalized PPO on train opponents}
        \begin{center}
            \begin{tabular}{|c|c|c|}
                \hline
                \textbf{Opponent}&{\textbf{Specialized NEAT}}&{\textbf{Generalized PPO}} \\
                \cline{2-4}
                \hline
                1 & 190.01 & 199.81 \\
                2 & 194.01 & 190.34 \\
                6 & 173.01 & 195.85 \\
                7 & 177.01 & 193.27 \\
                \hline
                \multicolumn{3}{l}{The generalized PPO was trained on the opponents \{1, 2, 6, 7\}}
            \end{tabular}
        \end{center}
    \end{table}

@@

    \begin{table}[htbp]
        \caption{Results}
        \begin{center}
            \begin{tabular}{|c|c|c|c|c|c|}
                \hline
                \textbf{Opponent}&\multicolumn{4}{|c|}{\textbf{PPO}}&\textbf{Specialized} \\
                \cline{2-5}
                & \textbf{\textit{Run 1}}& \textbf{\textit{Run 2}}& \multicolumn{2}{|c|}{\textbf{\textit{Run 3}}} & NEAT \\
                \cline{4-5}
                & 3000 ep & 3000 ep & 3000 ep & 1750 ep & \\
                & (best at training) &  &  & (best tested) & \\
                \hline
                1 & 198.41 & 199.07 & 199.81 & 199.61 & 190.01 \\
                \hline
                2 & 199.74 & 199.27 & 190.34 & 189.14 & 194.01 \\
                \hline
                3 & 58.94 & 46.67 & 70.27 &  85.01 & 180.01 \\
                \hline
                4 & 58.34 & 66.5 & 64.11 & 80.17 & 194.01 \\
                \hline
                5 & 172.61 & 165.65 & 174.61 & 158.83 & 194.01\\
                \hline
                6 & 195.73 & 196.95 & 195.85 & 194.37 & 173.01 \\
                \hline
                7 & 199.79 & 195.05 & 193.27 & 191.17 & 177.01 \\
                \hline
                8 & 122.17 & 145.49 & 141.89 & 140.61 & 186.01 \\
                \hline
                \textbf{\{1, 2, 6, 7\}} &  &  &  & & \\
                \textbf{harmonic mean} & 198.4 & 197.57 & 194.75 & 193.49 & 183.09 \\
                \hline
                \textbf{final} &  &  &  & & \\
                \textbf{harmonic mean} & 117.22 & 114.91 & 127.01 & 137.18 & 185.67\\
                \hline

            \end{tabular}
            \label{tab1}
        \end{center}
    \end{table}

    \subsection{Time Analysis}\label{subsec:time-analysis}
    Since the match time is relevant, the machine on which the games were run is also relevant.
    The tests were run on an i7 4750HQ processor.
    The training was done on 3 threads.
    All computations were done on the CPU\@.
    The median game time during training is 28 seconds.
    The average number of frames per game is 287.
